---
title: "Analysis of False Belief Task over Pre-training"
author: "Sean Trott, Cameron Jones, Pam Rivière"
date: "April 24, 2025"
output:
  pdf_document: 
       fig_caption: yes
       keep_md: yes
       keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "png")
```


```{r include=FALSE}
library(tidyverse)
library(lmtest)
library(forcats)
library(broom)
library(lme4)
library(viridis)
library(ggridges)
library(lmerTest)
library(ggrepel)
library(ggcorrplot)
library(dplyr)
library(ggplot2)

all_colors <- viridis::viridis(10, option = "mako")
my_colors <- all_colors[c(3, 5, 7)]  # Selecting specific colors from the palette
```

# Load LLM data

```{r}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/NLMs/epistemology/dev_tom/src/analysis")
# setwd("/Users/pamelariviere/Dropbox/Research/projects/dev_tom/src/analysis")

# Grab data for false belief task
directory_path <- "../../data/processed/fb_local/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_tmp <- bind_rows(csv_list) %>%
  mutate(model_shorthand = str_to_title(model_shorthand))
nrow(df_tmp)

df_tmp$tokens_seen_numeric <- as.numeric(sub("B", "", df_tmp$tokens_seen)) * 1e9

df_tmp = df_tmp %>%
  # filter(stage == "stage1") %>%
  mutate(model_id = paste(stage, "step", "-", step)) %>%
  mutate(tokens_seen_numeric_mod = tokens_seen_numeric + 1)


length(unique(df_tmp$model_id))
table(df_tmp$model_path)
table(df_tmp$model_shorthand)
table(df_tmp$tokens_seen_numeric)
table(df_tmp$step)

metadata <- read_csv("../../data/raw/metadata_models.csv")
# merge the metadata with the fb task df
df_all_models_fb <- df_tmp %>% left_join(metadata, by = "model_shorthand")


df_all_models_fb = df_all_models_fb %>%
  mutate(model_family = case_when(
    model_shorthand %in% c("Pythia 14m", "Pythia 1b", 
                           "Pythia 6.9b", "Pythia 12b") ~ "Pythia",
    model_shorthand %in% c("Olmo 2 1b", "Olmo 2 7b", "Olmo 2 13b") ~ "Olmo 2",
  ))


```


# FB Analysis

## Sensitivity to FB over time



## Accuracy metric


```{r}
df_all_models_fb = df_all_models_fb %>%
  mutate(correct = case_when(
    condition == "False Belief" & log_odds > 0 ~ TRUE,
    condition == "True Belief" & log_odds <= 0 ~ TRUE,
    TRUE ~ FALSE  # all other cases are incorrect
  ))


df_summ = df_all_models_fb %>%
  group_by(model_path, model_shorthand,
           step, tokens_seen_numeric, model_family, stage) %>%
  summarise(mean_accuracy = mean(correct)) %>%
  mutate(tokens_seen_numeric_mod = tokens_seen_numeric + 1)
df_summ %>%
  select(model_shorthand, mean_accuracy)
mean(df_all_models_fb$correct)

df_summ %>%
  ungroup() %>%
  arrange(desc(mean_accuracy)) %>%
  select(model_shorthand, mean_accuracy, step, stage) %>%
  head(5)

# Get the final step for each model and extract final accuracy
df_summ = df_all_models_fb %>%
  group_by(model_path, model_shorthand,
           step, tokens_seen_numeric, 
           model_family, stage, n_params_approx) %>%
  summarise(mean_accuracy = mean(correct)) %>%
  mutate(tokens_seen_numeric_mod = tokens_seen_numeric + 1)

final_accuracy <- df_summ %>%
  group_by(model_shorthand, model_family) %>%
  filter(step == max(step)) %>%  # Get the final step for each model
  ungroup()

final_accuracy <- df_summ %>%
  filter(!is.na(step), !is.na(model_shorthand)) %>%  # Remove NAs
  group_by(model_shorthand) %>%
  slice_max(step, n = 1, with_ties = FALSE) %>%  # One row per model, no ties
  ungroup()

# Plot accuracy at final step by model parameter size
ggplot(final_accuracy, aes(x = n_params_approx, y = mean_accuracy)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_text(aes(label = model_shorthand), vjust = -0.8, hjust = 0.5, size = 3) +
  geom_hline(yintercept = .5, linetype = "dashed",
             size = 1.2, alpha = .5) +
  scale_x_log10(labels = scales::comma_format()) +  # Log scale for params
  labs(
    title = "False Belief Task Accuracy at Final Step",
    x = "Number of Parameters (log scale)",
    y = "Accuracy"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.6),
        plot.subtitle = element_text(hjust = 0.6))

# Plot displays fb task accuracy for all steps, final step in red
ggplot(df_summ, aes(x = n_params_approx, y = mean_accuracy)) +
  geom_line(aes(group = model_shorthand), alpha = 0.3, color = "gray") +
  geom_point(alpha = 0.3, size = 1, color = "gray") +
  geom_point(data = final_accuracy, size = 4, 
             aes(color = model_family), 
             alpha = 0.8) +
  geom_hline(yintercept = .5, linetype = "dashed",
             size = 1.2, alpha = .5) +
  scale_color_manual(values = viridisLite::viridis(2, option = "mako", 
                                                  begin = 0.8, end = 0.15)) + 
  geom_text_repel(data = final_accuracy, aes(label = model_shorthand), size = 3) +
  scale_x_log10(labels = scales::comma_format()) +
  labs(
    title = "False Belief Task Accuracy: Progression and Final Step",
    subtitle = "Red points highlight accuracy at final pretraining step",
    x = "Number of Parameters (log scale)",
    y = "Accuracy"
  ) +
  theme_minimal()


df_summ %>%
  mutate(step_modded = step + 1) %>%
  ggplot(aes(x = step_modded,
             y = mean_accuracy,
             color = model_shorthand)) +
  #geom_point(size = 3,
  #           alpha = .7) +
  geom_hline(yintercept = .83,##TODO: Calculate from scratch
             linetype = "dotted", color = "red",
             size = 1.2, alpha = .8) + 
  geom_line(size = 1) +
  geom_hline(yintercept = .5, linetype = "dotted",
             size = 1.2, alpha = .5) +
  scale_x_log10() +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Step",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_bw() +
  scale_color_manual(values = viridisLite::viridis(7, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  facet_wrap(~stage)

## merge stepwise tokens data for Pythia models
tokens_seen_pythia <- read_csv("../../data/raw/pythia_tokens_seen.csv")
# merge the metadata with the fb task df
df_summ_tokenmerge <- df_summ %>%
  left_join(tokens_seen_pythia %>% select(model_shorthand, step, tokens_seen_from_df1 = tokens_seen_numeric), 
            by = c("model_shorthand", "step")) %>%
  mutate(tokens_seen_numeric = coalesce(tokens_seen_numeric, tokens_seen_from_df1)) %>%
  select(-tokens_seen_from_df1) %>%
  mutate(tokens_seen_numeric_mod = tokens_seen_numeric + 1)

df_summ_tokenmerge %>%
  ggplot(aes(x = tokens_seen_numeric_mod,
             y = mean_accuracy,
             color = model_shorthand)) +
  #geom_point(size = 3,
    #         alpha = .7) +
  geom_line(size = 1, alpha=0.7) +
  geom_hline(yintercept = .5, linetype = "dashed",
             size = 1.2, alpha = .5) +
  scale_x_log10(labels = scales::comma_format()) +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Tokens Seen",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_bw() +
  scale_color_manual(values = viridisLite::viridis(7, option = "mako", 
                                                  begin = 0.9, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  facet_wrap(~stage)



# Plot of Olmo stage1 and Pythias
df_summ_tokenmerge %>%
  filter(stage == "stage1" | is.na(stage)) %>%
  ggplot(aes(x = tokens_seen_numeric_mod,
             y = mean_accuracy,
             color = model_shorthand)) +
  geom_line(size = 1.5, alpha = 0.7) +
  geom_hline(yintercept = .5, linetype = "dashed",
             size = 1.2, alpha = .5) +
  scale_x_log10(labels = scales::comma_format()) +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Tokens Seen",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_bw() +
  scale_color_manual(values = viridisLite::viridis(7, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  labs(
    title = "False Belief Task Accuracy Over Pretraining",
    x = "Number of Tokens Seen (log scale)",
    y = "Accuracy"
  ) +
  theme(text = element_text(size = 15),
        legend.position="bottom") 


### How do model properties predict the probability of a correct response?
mod_full = glmer(data = df_all_models_fb,
                 correct ~ condition + knowledge_cue + 
                   log10(tokens_seen_numeric_mod) * stage +
                   (1 | start) +
                   (1 | model_shorthand),
                 family = binomial())
summary(mod_full)


### Plot coefficients
df_coef <- broom.mixed::tidy(mod_full, effects = "fixed") %>%
  mutate(term = forcats::fct_reorder(term, estimate))


df_coef %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x = term, y = estimate)) +
  geom_point() +
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error),
                width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  coord_flip() +
  labs(
    x = NULL, y = "Coefficient Estimate",
  ) +
  theme_minimal(base_size = 12)


# write.csv(df_summ, "../../data/processed/summaries/fb_summary.csv")
```

## Phase shift modeling

Here, we predict the probability of a correct response over tokens seen, focusing on stage 1.

### Using GAM

```{r}
library(mgcv)
library(dplyr)
library(purrr)

# Aggregate to get accuracy per checkpoint (helps GAM fit)
df_summary_fb <- df_all_models_fb %>%
  filter(stage == "stage1") %>%
  group_by(model_shorthand, 
           # condition, 
           # knowledge_cue,
           tokens_seen_numeric_mod) %>%
  summarise(acc = mean(correct), .groups = "drop") %>%
  mutate(log_tokens = log10(tokens_seen_numeric_mod))

# Fit a GAM for each model separately
fits <- df_summary_fb %>%
  group_split(model_shorthand) %>%
  set_names(map_chr(., ~ unique(.$model_shorthand))) %>%
  map(~ gam(acc ~ s(tokens_seen_numeric_mod, k = 10),
            data = ., family = binomial(link = "logit")))  # logistic is natural for accuracy

df_pred <- map2_df(fits, names(fits), function(mod, name) {
  newdat <- data.frame(tokens_seen_numeric_mod = seq(min(df_summary_fb$tokens_seen_numeric_mod),
                                        max(df_summary_fb$tokens_seen_numeric_mod),
                                        length.out = 200))
  newdat$acc_hat <- predict(mod, newdat, type = "response")
  newdat$model_shorthand <- name
  newdat
})

ggplot(df_summary_fb, aes(x = tokens_seen_numeric_mod, y = acc, color = model_shorthand)) +
  geom_point(alpha = 0.3) +
  geom_line(data = df_pred, aes(y = acc_hat), size = 1.2) +
  scale_color_manual(values = viridisLite::viridis(3, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  scale_x_log10(labels = scales::comma_format()) +
  labs(
    title = "Development of False Belief Task Performance",
    x = "Number of Tokens Seen (log scale)",
    y = "Accuracy"
  ) +
  theme_minimal()



### TODO: Identify inflection points
```


# Situation Model -- Attention Check Results


**TODO**: Unlike the original task, these aren't necessarily balanced; so differences in probability of start vs. end location, or which person, could also affect accuracy. Could check on this too.

```{r}
directory_path <- "../../data/processed/attn-checks-local/"

csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_olmo13_attn_check <- bind_rows(csv_list)
nrow(df_olmo13_attn_check)
table(df_olmo13_attn_check$model_path)


df_olmo13_attn_check$tokens_seen_numeric <- as.numeric(sub("B", "", df_olmo13_attn_check$tokens_seen)) * 1e9

## merge stepwise tokens data for Pythia models
tokens_seen_pythia <- read_csv("../../data/raw/pythia_tokens_seen.csv")
# merge the metadata with the fb task df
df_olmo13_attn_check <- df_olmo13_attn_check %>%
  left_join(tokens_seen_pythia %>% select(model_shorthand, step, tokens_seen_from_df1 = tokens_seen_numeric), 
            by = c("model_shorthand", "step")) %>%
  mutate(tokens_seen_numeric = coalesce(tokens_seen_numeric, tokens_seen_from_df1)) %>%
  select(-tokens_seen_from_df1) %>%
  mutate(tokens_seen_numeric_mod = tokens_seen_numeric + 1)

df_olmo13_attn_check = df_olmo13_attn_check %>%
  # filter(stage == "stage1") %>%
  mutate(model_id = paste(stage, "step", "-", step)) %>%
  mutate(tokens_seen_numeric_mod = tokens_seen_numeric + 1)


length(unique(df_olmo13_attn_check$model_id))
table(df_olmo13_attn_check$model_path)
table(df_olmo13_attn_check$model_shorthand)
table(df_olmo13_attn_check$tokens_seen_numeric)
table(df_olmo13_attn_check$step)


```

```{r}
df_olmo13_attn_check = df_olmo13_attn_check %>%
  mutate(correct = case_when(
    log_odds > 0 ~ TRUE,
    log_odds <= 0 ~ FALSE
  )) %>%
  mutate(q_label = case_when(
    question_id == 1 ~ "Item was first in {START}",
    question_id == 2 ~ "At end of story, item in {END}",
    question_id == 3 ~ "Original person was {X}",
    question_id == 4 ~ "Second person was {Y}"
  ))

## Accuracy of item start location
df_summ = df_olmo13_attn_check %>%
  group_by(model_path, model_shorthand, question_id,
           step, tokens_seen_numeric_mod, stage) %>%
  summarise(mean_accuracy = mean(correct)) %>%
  mutate(tokens_seen_numeric_mod = tokens_seen_numeric + 1)


df_summ %>%
  ungroup() %>%
  arrange(desc(mean_accuracy)) %>%
  select(model_path, step, mean_accuracy) %>%
  head(5)

df_summ = df_summ %>%
  mutate(q_label = case_when(
    question_id == 1 ~ "Item was first in {START}",
    question_id == 2 ~ "At end of story, item in {END}",
    question_id == 3 ~ "Original person was {X}",
    question_id == 4 ~ "Second person was {Y}"
  ))

df_summ %>%
  filter(stage == "stage1") %>%
  ggplot(aes(x = tokens_seen_numeric_mod,
             y = mean_accuracy,
             color = model_shorthand)) +
  geom_point(size = 3,
             alpha = .7) +
  geom_hline(yintercept = .5, linetype = "dotted",
             size = 1.2, alpha = .5) +
  scale_x_log10() +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Tokens Seen",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_bw() +
  scale_color_manual(values = viridisLite::viridis(4, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  facet_wrap(~reorder(q_label, question_id))

# plot same for the Pythias as well
df_summ %>%
  filter(grepl("Pythia", model_shorthand)) %>%
  ggplot(aes(x = tokens_seen_numeric_mod,
             y = mean_accuracy,
             color = model_shorthand)) +
  geom_point(size = 3,
             alpha = .7) +
  geom_hline(yintercept = .5, linetype = "dotted",
             size = 1.2, alpha = .5) +
  scale_x_log10(labels = scales::comma_format()) +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Tokens Seen",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_bw() +
  scale_color_manual(values = viridisLite::viridis(4, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  facet_wrap(~reorder(q_label, question_id))


### Final step
df_model_max = df_olmo13_attn_check %>%
  group_by(model_shorthand) %>%
  summarise(max_step = max(step, na.rm = TRUE))

df_olmo13_attn_check %>% 
  inner_join(df_model_max) %>%
  filter(step == max_step) %>%
  ggplot(aes(x = log_odds,
             y = factor(question_id),
             fill = model_shorthand)) +
  geom_density_ridges2(aes(height = ..density..), 
                       color=NA, 
                       scale=.85, 
                       # size=1, 
                       alpha = .8,
                       stat="density") +
  labs(x = "Log Odds (Start vs. End)",
       y = "",
       fill = "") +
  theme_minimal() +
  geom_vline(xintercept = 0, linetype = "dotted") +
  theme(
    legend.position = "bottom"
  ) + 
  theme(axis.title = element_text(size=rel(1.2)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        legend.title = element_text(size = rel(1.2)),
        strip.text.x = element_text(size = rel(1.2))) +
    scale_fill_manual(values = viridisLite::viridis(4, option = "mako", 
                                                    begin = 0.8, end = 0.15)) 


#df_olmo13_attn_check %>% 
#  inner_join(df_model_max) %>%
#  filter(step == max_step) %>%
#  group_by(model_shorthand, question_id) %>%
#  summarise(mean_accuracy = mean(correct),
#            step = mean(step))



#df_items = df_olmo13_attn_check %>% 
 # inner_join(df_model_max) %>%
 # filter(step == max_step) %>%
 # group_by(model_shorthand, item, question_id, q_label,
 #          correct_answer, distractor_answer) %>%
 # summarise(mean_accuracy = mean(correct))


#df_items %>%
#  filter(model_shorthand == "OLMO 2 13B") %>%
#  ggplot(aes(x = factor(item),
 #            y = mean_accuracy)) +
 ## geom_bar(stat = "identity") +
#  facet_wrap(~q_label) +
#  labs(x = "Scenario",
 #      y = "Accuracy") +
 # theme_minimal()

```



