---
title: "Analysis of False Belief Task over Pre-training"
author: "Sean Trott, Cameron Jones, Pam Rivière"
date: "April 24, 2025"
#output:
  #pdf_document: 
    #   fig_caption: yes
    #   keep_md: yes
    #   keep_tex: yes
    #   html_document:
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "png")

```


```{r include=FALSE}
library(tidyverse)
library(lmtest)
library(forcats)
library(broom)
library(lme4)
library(viridis)
library(ggridges)
library(lmerTest)
library(ggrepel)
library(ggcorrplot)
library(dplyr)
library(ggplot2)

```

# Load LLM False Belief Data

```{r}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/NLMs/epistemology/dev_tom/src/analysis")
# setwd("/Users/pamelariviere/Dropbox/Research/projects/dev_tom/src/analysis")

# Grab data for false belief task
directory_path <- "../../data/processed/fb_local/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_tmp <- bind_rows(csv_list) %>%
  mutate(model_shorthand = str_to_title(model_shorthand))
nrow(df_tmp)

# Create a column with numeric versions of the tokens seen for that step, using 
# the Olmo file naming convention
df_tmp$tokens_seen_numeric <- as.numeric(sub("B", "", df_tmp$tokens_seen)) * 1e9

# Create a column with a modified version of the numeric tokens seen, for plotting
# and analysis purposes (just +1)
df_tmp = df_tmp %>%
  # filter(stage == "stage1") %>%
  mutate(model_id = paste(stage, "step", "-", step)) %>%
  mutate(tokens_seen_numeric_mod = tokens_seen_numeric + 1)

## NOTE: Pythia model data contains a "main" checkpoint/step that appears as a NaN in the `step` column
## must change this to the actual value of the final step, 143000
df_tmp$step[is.na(df_tmp$step)] <- 143000 #hard-coded the final step 

# sort df columns by model name and step value
df_tmp <- df_tmp %>%
  arrange(model_shorthand, step)  # arrange in ascending order

metadata <- read_csv("../../data/raw/metadata_models.csv")
# merge the metadata with the fb task df
df_all_models_fb <- df_tmp %>% left_join(metadata, by = "model_shorthand")

tokens_seen_pythia <- read_csv("../../data/raw/pythia_tokens_seen.csv")
# merge stepwise tokens data for Pythia models
df_all_models_fb <- df_all_models_fb %>%
  left_join(tokens_seen_pythia %>% select(model_shorthand, step, tokens_seen_from_df1 = tokens_seen_numeric), 
            by = c("model_shorthand", "step")) %>%
  mutate(tokens_seen_numeric = coalesce(tokens_seen_numeric, tokens_seen_from_df1)) %>%
  select(-tokens_seen_from_df1) %>%
  mutate(tokens_seen_numeric_mod = tokens_seen_numeric + 1)


df_all_models_fb = df_all_models_fb %>%
  mutate(model_family = case_when(
    model_shorthand %in% c("Pythia 14m", "Pythia 1b", 
                           "Pythia 6.9b", "Pythia 12b") ~ "Pythia",
    model_shorthand %in% c("Olmo 2 1b", "Olmo 2 7b", "Olmo 2 13b") ~ "Olmo 2",
  ))

length(unique(df_tmp$model_id))
table(df_tmp$model_path)
table(df_tmp$model_shorthand)
table(df_tmp$tokens_seen_numeric)
table(df_tmp$step)

# How many unique checkpoints per model, per stage (in the case of stages)?
df_all_models_fb %>% 
  group_by(model_shorthand,stage) %>% 
  summarise(unique_steps = n_distinct(step))

df_all_models_fb %>% 
  group_by(model_shorthand,stage) %>% 
  summarise(unique_tokens = n_distinct(tokens_seen_numeric_mod))


# Tracking down duplicate tokens_seen_numeric_mod for different steps
# Find tokens_seen values that appear with multiple different steps
duplicates <- df_all_models_fb %>%
  group_by(model_shorthand, stage, tokens_seen_numeric_mod) %>%
  summarise(
    unique_steps = n_distinct(step),
    steps = list(unique(step)),
    .groups = "drop"
  ) %>%
  filter(unique_steps > 1)

# Get the actual rows where this occurs
problem_tokens <- duplicates$tokens_seen_numeric_mod

df_all_models_fb %>%
  filter(tokens_seen_numeric_mod %in% problem_tokens) %>%
  arrange(tokens_seen_numeric_mod, step)

tmp <- df_all_models_fb %>%
  filter(model_shorthand == "Olmo 2 7b") %>%
  filter(tokens_seen_numeric_mod %in% problem_tokens) %>% 
  arrange(tokens_seen_numeric_mod, step)

#For Olmo 2 7b --> see filenames for 
#   ....stage1_step600_3B, 
#.  ....stage1_step700_3B
#.  ....stage1_step850_4B
#.  ....stage1_step900_4B
# The tokens seen is duplicated for each of these steps, which is why you get 
# 73 unique steps, but 71 unique tokens seen for Olmo 2 7b, and this is just a 
# property of how the tokens seen appear on the 7b HuggingFace revisions
# this will mean that you should keep in mind which column name (`step` or 
# `tokens_seen_numeric_mod`) you want to group by for your analysis -- there are
# as many actually unique data points as there are number of `step` x number of conditions
 
```


# FB Analysis

## Sensitivity to FB over time

## Accuracy metric


```{r}
df_all_models_fb = df_all_models_fb %>%
  mutate(correct = case_when(
    condition == "False Belief" & log_odds > 0 ~ TRUE,
    condition == "True Belief" & log_odds <= 0 ~ TRUE,
    TRUE ~ FALSE  # all other cases are incorrect
  ))

df_summ_fb = df_all_models_fb %>%
  group_by(model_path, model_shorthand,
           step, tokens_seen_numeric, model_family, stage) %>%
  summarise(mean_accuracy = mean(correct)) %>%
  mutate(tokens_seen_numeric_mod = tokens_seen_numeric + 1)
df_summ_fb %>%
  select(model_shorthand, mean_accuracy)
mean(df_all_models_fb$correct)

df_summ_fb %>%
  ungroup() %>%
  arrange(desc(mean_accuracy)) %>%
  select(model_shorthand, mean_accuracy, step, stage) %>%
  head(5)

# Get the final step for each model and extract final accuracy
df_summ_fb = df_all_models_fb %>%
  group_by(model_path, model_shorthand,
           step, tokens_seen_numeric, 
           model_family, stage, n_params_approx) %>%
  summarise(mean_accuracy = mean(correct)) %>%
  mutate(tokens_seen_numeric_mod = tokens_seen_numeric + 1)

final_accuracy <- df_summ_fb %>%
  group_by(model_shorthand, model_family) %>%
  filter(step == max(step)) %>%  # Get the final step for each model
  ungroup()

final_accuracy <- df_summ_fb %>%
  filter(!is.na(step), !is.na(model_shorthand)) %>%  # Remove NAs
  group_by(model_shorthand) %>%
  slice_max(step, n = 1, with_ties = FALSE) %>%  # One row per model, no ties
  ungroup()

# Plot accuracy at final step by model parameter size
ggplot(final_accuracy, aes(x = n_params_approx, y = mean_accuracy)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_text(aes(label = model_shorthand), vjust = -0.8, hjust = 0.5, size = 3) +
  geom_hline(yintercept = .5, linetype = "dashed",
             size = 1.2, alpha = .5) +
  scale_x_log10(labels = scales::comma_format()) +  # Log scale for params
  labs(
    title = "False Belief Task Accuracy at Final Step",
    x = "Number of Parameters (log scale)",
    y = "Accuracy"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.6),
        plot.subtitle = element_text(hjust = 0.6))

# PAPER FIGURE: Plot displays fb task accuracy for all steps, final step in red
ggplot(df_summ_fb, aes(x = n_params_approx, y = mean_accuracy)) +
  geom_line(aes(group = model_shorthand), alpha = 0.3, color = "gray") +
  geom_point(alpha = 0.3, size = 1, color = "gray") +
  geom_point(data = final_accuracy, size = 4, 
             aes(color = model_family), 
             alpha = 0.8) +
  geom_hline(yintercept = .5, linetype = "dashed",
             size = 1.2, alpha = .5) +
  scale_color_manual(values = viridisLite::viridis(2, option = "mako", 
                                                  begin = 0.8, end = 0.15)) + 
  geom_text_repel(data = final_accuracy, aes(label = model_shorthand), size = 3) +
  scale_x_log10(labels = scales::comma_format()) +
  labs(
    title = "False Belief Task Accuracy: Progression and Final Step",
    subtitle = "Red points highlight accuracy at final pretraining step",
    x = "Number of Parameters (log scale)",
    y = "Accuracy"
  ) +
  theme_minimal()

## PAPER FIGURE: filter out the Pythia-14M, since it doesn't add much? 
final_accuracy_exclude14m <- final_accuracy %>%
  filter(model_shorthand != "Pythia 14m")
df_summ_fb %>%
  filter(model_shorthand != "Pythia 14m") %>%
  ggplot(aes(x = n_params_approx, y = mean_accuracy)) + 
  geom_line(aes(group = model_shorthand), alpha = 0.3, color = "gray") +
  geom_point(alpha = 0.3, size = 1, color = "gray") +
  geom_point(data = final_accuracy_exclude14m, size = 4, 
             aes(color = model_family), 
             alpha = 0.8) +
  geom_hline(yintercept = .5, linetype = "dashed",
             size = 1.2, alpha = .5) +
  scale_color_manual(values = viridisLite::viridis(2, option = "mako", 
                                                  begin = 0.8, end = 0.15)) + 
  geom_text_repel(data = final_accuracy_exclude14m, aes(label = model_shorthand), size = 3) +
  scale_x_log10(labels = scales::comma_format()) +
  labs(
    title = "False Belief Task Accuracy: Progression and Final Step",
    subtitle = "Colored points highlight accuracy at final pretraining step",
    x = "Number of Parameters (log scale)",
    y = "Accuracy"
  ) +
  theme_minimal()


df_summ_fb %>%
  mutate(step_modded = step + 1) %>%
  ggplot(aes(x = step_modded,
             y = mean_accuracy,
             color = model_shorthand)) +
  #geom_point(size = 3,
  #           alpha = .7) +
  geom_hline(yintercept = .83,##TODO: Calculate from scratch
             linetype = "dotted", color = "red",
             size = 1.2, alpha = .8) + 
  geom_line(size = 1) +
  geom_hline(yintercept = .5, linetype = "dotted",
             size = 1.2, alpha = .5) +
  scale_x_log10() +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Step",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_bw() +
  scale_color_manual(values = viridisLite::viridis(7, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  facet_wrap(~stage)

df_summ_fb %>%
  ggplot(aes(x = tokens_seen_numeric_mod,
             y = mean_accuracy,
             color = model_shorthand)) +
  #geom_point(size = 3,
    #         alpha = .7) +
  geom_line(size = 1, alpha=0.7) +
  geom_hline(yintercept = .5, linetype = "dashed",
             size = 1.2, alpha = .5) +
  scale_x_log10(labels = scales::comma_format()) +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Tokens Seen",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_bw() +
  scale_color_manual(values = viridisLite::viridis(7, option = "mako", 
                                                  begin = 0.9, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  facet_wrap(~stage)

# PAPER FIGURE: Plot of Olmo stage1 and Pythias
df_summ_fb %>%
  filter(stage == "stage1" | is.na(stage)) %>%
  ggplot(aes(x = tokens_seen_numeric_mod,
             y = mean_accuracy,
             color = model_shorthand)) +
  geom_line(size = 1.5, alpha = 0.8) +
  geom_hline(yintercept = .5, linetype = "dashed",
             size = 1.2, alpha = .5) +
  scale_x_log10(labels = scales::comma_format()) +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  #scale_y_continuous(limits = c(0.25, 0.75)) +
  labs(x = "Tokens Seen",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_bw() +
  scale_color_manual(values = viridisLite::viridis(7, option = "mako", 
                                                  begin = 0.8, end = 0.1)) +
  labs(
    title = "False Belief Task Accuracy Over Pretraining",
    x = "Number of Tokens Seen (log scale)",
    y = "Accuracy"
  ) +
  theme_minimal() + 
  theme(text = element_text(size = 15),
        legend.position="bottom") 


# Filter for only stage1 olmo data and pythia models (e.g. exclude olmo stage2)
olmo_and_pythia <- df_all_models_fb %>%
  filter(stage == "stage1" | is.na(stage)) 

  
### PAPER LMER: How do model properties predict the probability of a correct response?
mod_full = glmer(data = olmo_and_pythia,
                 correct ~ condition + knowledge_cue + 
                   log10(tokens_seen_numeric_mod) +
                   (1 | start) +
                   (1 | model_shorthand),
                 family = binomial())
summary(mod_full)


### Plot coefficients
df_coef <- broom.mixed::tidy(mod_full, effects = "fixed") %>%
  mutate(term = forcats::fct_reorder(term, estimate))


df_coef %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x = term, y = estimate)) +
  geom_point() +
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error),
                width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  coord_flip() +
  labs(
    x = NULL, y = "Coefficient Estimate",
  ) +
  theme_minimal(base_size = 12)


# write.csv(df_summ, "../../data/processed/summaries/fb_summary.csv")
```


```{r}
# Find the tokens seen for Olmo 2 13b as the ramp in accuracy is taking place

# Filter for Olmo 2 13b data
olmo_data <- df_summ_fb[df_summ_fb$model_shorthand == "Olmo 2 13b", ]

# Calculate the median accuracy for this model
median_accuracy_across_training <- median(olmo_data$mean_accuracy)

# Find the row(s) where accuracy is closest to the median
# Calculate absolute difference from median
olmo_data$diff_from_median <- abs(olmo_data$mean_accuracy - median_accuracy_across_training)

# Find the tokens_seen value where accuracy is closest to median
closest_row <- olmo_data[which.min(olmo_data$diff_from_median), ]
tokens_at_median <- closest_row$tokens_seen_numeric_mod

# Display results
print(paste("Median accuracy:", median_accuracy_across_training))
print(paste("Tokens seen at median accuracy:", tokens_at_median))
print(paste("Actual accuracy at that point:", closest_row$mean_accuracy))

# visualize to see if you've found the intended checkpoints

df_summ_fb %>%
  filter(stage == "stage1" | is.na(stage)) %>%
  ggplot(aes(x = tokens_seen_numeric_mod,
             y = mean_accuracy,
             color = model_shorthand)) +
  geom_line(size = 1.5, alpha = 0.8) +
  geom_vline(xintercept = closest_row$tokens_seen_numeric_mod, alpha = 0.8) +
  geom_hline(yintercept = .5, linetype = "dashed",
             size = 1.2, alpha = .5) +
  scale_x_log10(labels = scales::comma_format()) +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  #scale_y_continuous(limits = c(0.25, 0.75)) +
  labs(x = "Tokens Seen",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_bw() +
  scale_color_manual(values = viridisLite::viridis(7, option = "mako", 
                                                  begin = 0.8, end = 0.1)) +
  labs(
    title = "False Belief Task Accuracy Over Pretraining",
    x = "Number of Tokens Seen (log scale)",
    y = "Accuracy"
  ) +
  theme_minimal() + 
  theme(text = element_text(size = 15),
        legend.position="bottom") 

```

```{r}
# Find the tokens seen for Olmo 2 7b as the ramp in accuracy is taking place

# Filter for Olmo 2 13b data
olmo_data <- df_summ_fb[df_summ_fb$model_shorthand == "Olmo 2 7b", ]

# Calculate the median accuracy for this model
median_accuracy_across_training <- median(olmo_data$mean_accuracy)

# ...and the 75th percentile accuracy
pctile_75_across_training <- quantile(olmo_data$mean_accuracy, probs = 0.75)  # 75th percentile

# ...and the max accuracy
max_across_training <- max(olmo_data$mean_accuracy)

# Find the row(s) where accuracy is closest to the median
# Calculate absolute difference from median
olmo_data$diff_from_median <- abs(olmo_data$mean_accuracy - median_accuracy_across_training)

# ...also for 75th percentile
olmo_data$diff_from_75th <- abs(olmo_data$mean_accuracy - pctile_75_across_training)

# ...also for max accuracy
olmo_data$diff_from_max <- abs(olmo_data$mean_accuracy - max_across_training)

# Find the tokens_seen value where accuracy is closest to median
closest_row <- olmo_data[which.min(olmo_data$diff_from_median), ]
tokens_at_median <- closest_row$tokens_seen_numeric_mod

# ...also for 75th percentile
closest_row_75th <- olmo_data[which.min(olmo_data$diff_from_75th), ]
tokens_at_75th <- closest_row_75th$tokens_seen_numeric_mod

# ...also for max accuracy
closest_row_max <- olmo_data[which.min(olmo_data$diff_from_max), ]
tokens_at_max <- closest_row_max$tokens_seen_numeric_mod

# Display results
print(paste("Median accuracy:", median_accuracy_across_training))
print(paste("Tokens seen at median accuracy:", tokens_at_median))
print(paste("Actual accuracy at that point:", closest_row$mean_accuracy))

# ...also for 75th percentile
print(paste("75th percentile accuracy:", pctile_75_across_training))
print(paste("Tokens seen at 75th percentile accuracy:", tokens_at_75th))
print(paste("Actual accuracy at that point:", closest_row_75th$mean_accuracy))

# ...also for max accuracy
print(paste("max  accuracy:", max_across_training))
print(paste("Tokens seen at max accuracy:", tokens_at_max))
print(paste("Actual accuracy at that point:", closest_row_max$mean_accuracy))

# visualize to see if you've found the intended checkpoints

df_summ_fb %>%
  filter(stage == "stage1" | is.na(stage)) %>%
  ggplot(aes(x = tokens_seen_numeric_mod,
             y = mean_accuracy,
             color = model_shorthand)) +
  geom_line(size = 1.5, alpha = 0.8) +
  geom_vline(xintercept = closest_row$tokens_seen_numeric_mod, alpha = 0.8, linetype = "dashed") +
  geom_vline(xintercept = closest_row_75th$tokens_seen_numeric_mod, alpha = 0.8, linetype = "dashed") + 
  geom_vline(xintercept = closest_row_max$tokens_seen_numeric_mod, alpha = 0.8, linetype = "dashed", color = "red") +
  geom_hline(yintercept = .5, linetype = "dashed",
             size = 1.2, alpha = .5) +
  scale_x_log10(labels = scales::comma_format()) +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  #scale_y_continuous(limits = c(0.25, 0.75)) +
  labs(x = "Tokens Seen",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_bw() +
  scale_color_manual(values = viridisLite::viridis(7, option = "mako", 
                                                  begin = 0.8, end = 0.1)) +
  labs(
    title = "False Belief Task Accuracy Over Pretraining",
    x = "Number of Tokens Seen (log scale)",
    y = "Accuracy"
  ) +
  theme_minimal() + 
  theme(text = element_text(size = 15),
        legend.position="bottom") 

```
## Phase shift modeling

Here, we predict the probability of a correct response over tokens seen, focusing on stage 1.

### Using GAM

```{r}
library(mgcv)
library(dplyr)
library(purrr)

# Aggregate to get accuracy per checkpoint (helps GAM fit) for just Olmo stage1 & pythia
df_summary_fb <- olmo_and_pythia %>%
  filter(stage == "stage1" | is.na(stage)) %>%
  group_by(model_shorthand, 
           condition, 
           knowledge_cue,
           step,
           tokens_seen_numeric_mod) %>%
  summarise(acc = mean(correct), .groups = "drop") %>%
  mutate(log_tokens = log10(tokens_seen_numeric_mod))


# Fit a GAM for each model separately
fits <- df_summary_fb %>%
  group_split(model_shorthand) %>%
  set_names(map_chr(., ~ unique(.$model_shorthand))) %>%
  map(~ gam(acc ~ s(tokens_seen_numeric_mod, k = 10),
            data = ., family = binomial(link = "logit")))  # logistic is natural for accuracy

df_pred <- map2_df(fits, names(fits), function(mod, name) {
  newdat <- data.frame(tokens_seen_numeric_mod = seq(min(df_summary_fb$tokens_seen_numeric_mod),
                                        max(df_summary_fb$tokens_seen_numeric_mod),
                                        length.out = 200))
  newdat$acc_hat <- predict(mod, newdat, type = "response")
  newdat$model_shorthand <- name
  newdat
})

# PAPER FIGURE: plot of GAM fits for each model
ggplot(df_summary_fb, aes(x = tokens_seen_numeric_mod, y = acc, color = model_shorthand)) +
  geom_point(alpha = 0.3) +
  geom_line(data = df_pred, aes(y = acc_hat), size = 1.2) +
  scale_color_manual(values = viridisLite::viridis(7, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  scale_x_log10(labels = scales::comma_format()) +
  labs(
    title = "Development of False Belief Task Performance",
    x = "Number of Tokens Seen (log scale)",
    y = "Accuracy"
  ) +
  theme_minimal()



### TODO: Identify inflection points
```


# Situation Model -- Attention Check Results


**TODO**: Unlike the original task, these aren't necessarily balanced; so differences in probability of start vs. end location, or which person, could also affect accuracy. Could check on this too.


```{r}
directory_path <- "../../data/processed/attn-checks-local/"

csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_attn_check <- bind_rows(csv_list)
nrow(df_attn_check)
table(df_attn_check$model_path)

df_attn_check$tokens_seen_numeric <- as.numeric(sub("B", "", df_attn_check$tokens_seen)) * 1e9

df_attn_check = df_attn_check %>%
  mutate(model_id = paste(stage, "step", "-", step)) %>%
  mutate(tokens_seen_numeric_mod = tokens_seen_numeric + 1)

## NOTE: Pythia model data contains a "main" checkpoint/step that appears as a NaN in the `step` column
## must change this to the actual value of the final step, 143000
df_attn_check$step[is.na(df_attn_check$step)] <- 143000 #hard-coded the final step 

# sort df columns by model name and step value
df_attn_check <- df_attn_check %>%
  arrange(model_shorthand, step)  # arrange in ascending order

metadata <- read_csv("../../data/raw/metadata_models.csv")
# merge the metadata with the fb task df
df_attn_check <- df_attn_check %>% left_join(metadata, by = "model_shorthand")


# Change the model shorthand naming convention to match that in the df_attn_check 
# dataframe!
tokens_seen_pythia <- tokens_seen_pythia %>%
  mutate(model_shorthand = str_replace(model_shorthand, "b", "B"))

# merge stepwise tokens data for Pythia models
df_attn_check <- df_attn_check %>%
  left_join(tokens_seen_pythia %>% select(model_shorthand, step, tokens_seen_from_df1 = tokens_seen_numeric), 
            by = c("model_shorthand", "step")) %>%
  mutate(tokens_seen_numeric = coalesce(tokens_seen_numeric, tokens_seen_from_df1)) %>%
  select(-tokens_seen_from_df1) %>%
  mutate(tokens_seen_numeric_mod = tokens_seen_numeric + 1)

length(unique(df_attn_check$model_id))
table(df_attn_check$model_path)
table(df_attn_check$model_shorthand)
table(df_attn_check$tokens_seen_numeric)
table(df_attn_check$step)


```

```{r}
df_attn_check = df_attn_check %>%
  mutate(correct = case_when(
    log_odds > 0 ~ TRUE,
    log_odds <= 0 ~ FALSE
  )) %>%
  mutate(q_label = case_when(
    question_id == 1 ~ "Item was first in {START}",
    question_id == 2 ~ "At end of story, item in {END}",
    question_id == 3 ~ "Original person was {X}",
    question_id == 4 ~ "Second person was {Y}"
  ))

## Accuracy of item start location
df_summ_attn = df_attn_check %>%
  group_by(model_path, model_shorthand, question_id,
           step, tokens_seen_numeric_mod, stage) %>%
  summarise(mean_accuracy = mean(correct)) 


df_summ_attn %>%
  ungroup() %>%
  arrange(desc(mean_accuracy)) %>%
  select(model_path, step, mean_accuracy) %>%
  head(5)

df_summ_attn = df_summ_attn %>%
  mutate(q_label = case_when(
    question_id == 1 ~ "Item was first in {START}",
    question_id == 2 ~ "At end of story, item in {END}",
    question_id == 3 ~ "Original person was {X}",
    question_id == 4 ~ "Second person was {Y}"
  ))

# PAPER FIGURE - attention check results per question
df_summ_attn %>%
  filter(stage == "stage1" | is.na(stage)) %>%
  ggplot(aes(x = tokens_seen_numeric_mod,
             y = mean_accuracy,
             color = model_shorthand)) +
  #geom_line(size = 1,
  #           alpha = .5)+
  geom_point(size = 3,
             alpha = .7) +
  geom_hline(yintercept = .5, linetype = "dotted",
             size = 1.2, alpha = .5) +
  scale_x_log10() +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Development of Situation Model", 
       x = "Tokens Seen",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_bw() +
  scale_color_manual(values = viridisLite::viridis(7, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  facet_wrap(~reorder(q_label, question_id))

# PAPER FIGURE: plot same for the Pythias as well
df_summ_attn %>%
  filter(grepl("Pythia", model_shorthand)) %>%
  ggplot(aes(x = tokens_seen_numeric_mod,
             y = mean_accuracy,
             color = model_shorthand)) +
  geom_point(size = 3,
             alpha = .7) +
  geom_hline(yintercept = .5, linetype = "dotted",
             size = 1.2, alpha = .5) +
  scale_x_log10(labels = scales::comma_format()) +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Development of Situation Model",
       x = "Tokens Seen",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_bw() +
  scale_color_manual(values = viridisLite::viridis(4, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  facet_wrap(~reorder(q_label, question_id))


### Final step
df_model_max = df_attn_check %>%
  group_by(model_shorthand) %>%
  summarise(max_step = max(step, na.rm = TRUE))

df_attn_check %>% 
  inner_join(df_model_max) %>%
  filter(step == max_step) %>%
  ggplot(aes(x = log_odds,
             y = factor(question_id),
             fill = model_shorthand)) +
  geom_density_ridges2(aes(height = ..density..), 
                       color=NA, 
                       scale=.85, 
                       # size=1, 
                       alpha = .6,
                       stat="density") +
  labs(x = "Log Odds (Start vs. End)",
       y = "",
       fill = "") +
  theme_minimal() +
  geom_vline(xintercept = 0, linetype = "dotted") +
  theme(
    legend.position = "bottom"
  ) + 
  theme(axis.title = element_text(size=rel(1.2)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        legend.title = element_text(size = rel(1.2)),
        strip.text.x = element_text(size = rel(1.2))) +
    scale_fill_manual(values = viridisLite::viridis(7, option = "mako", 
                                                    begin = 0.8, end = 0.15)) 


#df_olmo13_attn_check %>% 
#  inner_join(df_model_max) %>%
#  filter(step == max_step) %>%
#  group_by(model_shorthand, question_id) %>%
#  summarise(mean_accuracy = mean(correct),
#            step = mean(step))



#df_items = df_olmo13_attn_check %>% 
 # inner_join(df_model_max) %>%
 # filter(step == max_step) %>%
 # group_by(model_shorthand, item, question_id, q_label,
 #          correct_answer, distractor_answer) %>%
 # summarise(mean_accuracy = mean(correct))


#df_items %>%
#  filter(model_shorthand == "OLMO 2 13B") %>%
#  ggplot(aes(x = factor(item),
 #            y = mean_accuracy)) +
 ## geom_bar(stat = "identity") +
#  facet_wrap(~q_label) +
#  labs(x = "Scenario",
 #      y = "Accuracy") +
 # theme_minimal()

```

```{r}
# Want to plot:
# 1. overall Situation Model accuracy superimposed over False Belief Task accuracy
# 2. individual situation model items accuracy over False Belief Task accuracy

## ====== ====== ====== ====== ====== 
## ====== FB + ATTN CHECK 1 ======
## ====== ====== ====== ====== ======
# Fully join false belief and filtered attention check dataframes, then pivot into tidy data
# creating `task_type` column 
fb_and_attn_q1 <- bind_rows(
  df_summ_fb %>% mutate(task_type = "false belief"),
  df_summ_attn %>%
    filter(question_id == 1) %>%
    mutate(task_type = "situation model 1")) %>%
    mutate(model_shorthand = str_replace(model_shorthand, "B$", "b")) %>%
    mutate(model_shorthand = str_replace(model_shorthand, "OLMO", "Olmo"))

# Define your desired order
model_order <- c("Pythia 14m", "Pythia 1b", "Pythia 6.9b", "Pythia 12b", 
                 "Olmo 2 1b", "Olmo 2 7b", "Olmo 2 13b")

# Convert to factor with desired order
fb_and_attn_q1 <- fb_and_attn_q1 %>%
  mutate(model_shorthand = factor(model_shorthand, levels = model_order))

# Filter for individual attention check questions
# Extract the unique question label
subtitle_text_1 <- df_summ_attn %>%
  filter(question_id == 1) %>%
  filter(stage == "stage1" | is.na(stage)) %>%
  pull(q_label) %>%
  unique()

# PAPER APPENDIX FIG: Plot accuracy for attention check 1 and false belief by tokens seen
# PAPER APPENDIX FIG: Plot accuracy for attention check 1 and false belief by tokens seen
fb_and_attn_q1 %>%
  filter(stage == "stage1" | is.na(stage)) %>%
  ggplot(aes(x = tokens_seen_numeric_mod,
             y = mean_accuracy,
             color = task_type)) +
  geom_line(size = 1,
             alpha = .5) +
  scale_x_log10() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Development of Situation Model",
       subtitle = subtitle_text_1,
       x = "Tokens Seen",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_bw() +
  scale_color_manual(values = viridisLite::viridis(2, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") + 
  facet_wrap(~model_shorthand)


fb_and_attn_q1 %>%
  filter(model_shorthand == "Olmo 2 13b") %>%
  filter(stage == "stage1" | is.na(stage)) %>%
  ggplot(aes(x = tokens_seen_numeric_mod,
             y = mean_accuracy,
             color = task_type)) +
  geom_line(size = 1,
             alpha = .5) +
  scale_x_log10() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
       subtitle = subtitle_text_1,
       x = "Tokens Seen",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_minimal() +
  scale_color_manual(values = viridisLite::viridis(2, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") 


## ====== ====== ====== ====== ====== 
## ====== FB + ATTN CHECK 2 ======
## ====== ====== ====== ====== ======
# Fully join false belief and filtered attention check dataframes, then pivot into tidy data
# creating `task_type` column 
fb_and_attn_q2 <- bind_rows(
  df_summ_fb %>% mutate(task_type = "false belief"),
  df_summ_attn %>%
    filter(question_id == 2) %>%
    mutate(task_type = "situation model 2")) %>%
    mutate(model_shorthand = str_replace(model_shorthand, "B$", "b")) %>%
    mutate(model_shorthand = str_replace(model_shorthand, "OLMO", "Olmo"))

# Convert to factor with desired order
fb_and_attn_q2 <- fb_and_attn_q2 %>%
  mutate(model_shorthand = factor(model_shorthand, levels = model_order))

# Filter for individual attention check questions
# Extract the unique question label
subtitle_text_2 <- df_summ_attn %>%
  filter(question_id == 2) %>%
  filter(stage == "stage1" | is.na(stage)) %>%
  pull(q_label) %>%
  unique()

# PAPER APPENDIX FIG: Plot accuracy for attention check 2 and false belief by tokens seen
fb_and_attn_q2 %>%
  filter(stage == "stage1" | is.na(stage)) %>%
  ggplot(aes(x = tokens_seen_numeric_mod,
             y = mean_accuracy,
             color = task_type)) +
  geom_line(size = 1,
             alpha = .5) +
  scale_x_log10() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Development of Situation Model",
       subtitle = subtitle_text_2,
       x = "Tokens Seen",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_bw() +
  scale_color_manual(values = viridisLite::viridis(2, option = "mako", 
                                                  begin = 0.7, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") + 
  facet_wrap(~model_shorthand)

# PAPER APPENDIX FIG: Plot accuracy for attention check 2 and false belief by tokens seen
fb_and_attn_q2 %>%
  filter(model_shorthand == "Olmo 2 13b") %>%
  filter(stage == "stage1" | is.na(stage)) %>%
  ggplot(aes(x = tokens_seen_numeric_mod,
             y = mean_accuracy,
             color = task_type)) +
  geom_line(size = 1,
             alpha = .5) +
  scale_x_log10() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(subtitle = subtitle_text_2,
       x = "Tokens Seen",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_minimal() +
  scale_color_manual(values = viridisLite::viridis(2, option = "mako", 
                                                  begin = 0.7, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") 


## ====== ====== ====== ====== ====== 
## ====== FB + ATTN CHECK 3 ======
## ====== ====== ====== ====== ======
# Fully join false belief and filtered attention check dataframes, then pivot into tidy data
# creating `task_type` column 
fb_and_attn_q3 <- bind_rows(
  df_summ_fb %>% mutate(task_type = "false belief"),
  df_summ_attn %>%
    filter(question_id == 3) %>%
    mutate(task_type = "situation model 3")) %>%
    mutate(model_shorthand = str_replace(model_shorthand, "B$", "b")) %>%
    mutate(model_shorthand = str_replace(model_shorthand, "OLMO", "Olmo"))

# Convert to factor with desired order
fb_and_attn_q3 <- fb_and_attn_q3 %>%
  mutate(model_shorthand = factor(model_shorthand, levels = model_order))

# Filter for individual attention check questions
# Extract the unique question label
subtitle_text_3 <- df_summ_attn %>%
  filter(question_id == 3) %>%
  filter(stage == "stage1" | is.na(stage)) %>%
  pull(q_label) %>%
  unique()

# PAPER APPENDIX FIG: Plot accuracy for attention check 3 and false belief by tokens seen
fb_and_attn_q3 %>%
  filter(stage == "stage1" | is.na(stage)) %>%
  ggplot(aes(x = tokens_seen_numeric_mod,
             y = mean_accuracy,
             color = task_type)) +
  geom_line(size = 1,
             alpha = .5) +
  scale_x_log10() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Development of Situation Model",
       subtitle = subtitle_text_3,
       x = "Tokens Seen",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_bw() +
  scale_color_manual(values = viridisLite::viridis(2, option = "mako", 
                                                  begin = 0.7, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") + 
  facet_wrap(~model_shorthand)

 

fb_and_attn_q3 %>%
  filter(model_shorthand == "Olmo 2 13b") %>%
  ggplot(aes(x = tokens_seen_numeric_mod,
             y = mean_accuracy,
             color = task_type)) +
  geom_line(size = 1,
             alpha = .5) +
  scale_x_log10() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(subtitle = subtitle_text_3,
       x = "Tokens Seen",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_minimal() +
  scale_color_manual(values = viridisLite::viridis(2, option = "mako", 
                                                  begin = 0.7, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") 
 
## ====== ====== ====== ====== ====== 
## ====== FB + ATTN CHECK 4 ======
## ====== ====== ====== ====== ======
# Fully join false belief and filtered attention check dataframes, then pivot into tidy data
# creating `task_type` column 
fb_and_attn_q4 <- bind_rows(
  df_summ_fb %>% mutate(task_type = "false belief"),
  df_summ_attn %>%
    filter(question_id == 4) %>%
    mutate(task_type = "situation model 4")) %>%
    mutate(model_shorthand = str_replace(model_shorthand, "B$", "b")) %>%
    mutate(model_shorthand = str_replace(model_shorthand, "OLMO", "Olmo"))

# Convert to factor with desired order
fb_and_attn_q4 <- fb_and_attn_q4 %>%
  mutate(model_shorthand = factor(model_shorthand, levels = model_order))

# Filter for individual attention check questions
# Extract the unique question label
subtitle_text_4 <- df_summ_attn %>%
  filter(question_id == 4) %>%
  filter(stage == "stage1" | is.na(stage)) %>%
  pull(q_label) %>%
  unique()

# PAPER APPENDIX FIG: Plot accuracy for attention check 3 and false belief by tokens seen
fb_and_attn_q4 %>%
  filter(stage == "stage1" | is.na(stage)) %>%
  ggplot(aes(x = tokens_seen_numeric_mod,
             y = mean_accuracy,
             color = task_type)) +
  geom_line(size = 1,
             alpha = .5) +
  scale_x_log10() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(subtitle = subtitle_text_4,
       x = "Tokens Seen",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_bw() +
  scale_color_manual(values = viridisLite::viridis(2, option = "mako", 
                                                  begin = 0.7, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") + 
  facet_wrap(~model_shorthand)


fb_and_attn_q4 %>%
  filter(model_shorthand == "Olmo 2 13b") %>%
  ggplot(aes(x = tokens_seen_numeric_mod,
             y = mean_accuracy,
             color = task_type)) +
  geom_line(size = 1,
             alpha = .5) +
  scale_x_log10() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Development of Situation Model",
       subtitle = subtitle_text_4,
       x = "Tokens Seen",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_minimal() +
  scale_color_manual(values = viridisLite::viridis(2, option = "mako", 
                                                  begin = 0.7, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") 

## ====== ====== ====== ====== ====== 
## ====== FB + ALL ATTN CHECKS ======
## ====== ====== ====== ====== ======

# Concatenate the individual attention check dataframes, keeping only one instance
# of the rows corresponding to false belief task accuracies
fb_and_all_attn <- bind_rows(
  fb_and_attn_q1,
  fb_and_attn_q2 %>% filter(task_type == "situation model 2"),
  fb_and_attn_q3 %>% filter(task_type == "situation model 3"),
  fb_and_attn_q4 %>% filter(task_type == "situation model 4")
)

  
situation_model_avg <- fb_and_all_attn %>%
  filter(task_type != "false belief task") %>%
  group_by(model_family, model_shorthand, step, tokens_seen_numeric, tokens_seen_numeric_mod) %>%
  summarise(mean_accuracy = mean(mean_accuracy, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(task_type = "situation model")

# Now combine this new df with the false belief task df
fb_and_all_attn <- bind_rows(
  fb_and_attn_q1 %>% filter(task_type == "false belief"),
  situation_model_avg
)

# Convert to factor with desired order
fb_and_all_attn <- fb_and_all_attn %>%
  mutate(model_shorthand = factor(model_shorthand, levels = model_order))

## PAPER FIGURE: FB & OVERALL SITUATION MODEL
fb_and_all_attn %>%
  filter(stage == "stage1" | is.na(stage)) %>%
  ggplot(aes(x = tokens_seen_numeric_mod,
             y = mean_accuracy,
             color = task_type)) + 
   geom_line(size = 1,
             alpha = .5) + 
  scale_x_log10() + 
  scale_y_continuous(limits = c(0, 1)) + 
  labs(title = "Development of Situation Model",
       x = "Tokens Seen",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_bw() +
  scale_color_manual(values = viridisLite::viridis(2, option = "mako", 
                                                  begin = 0.7, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") + 
  facet_wrap(~model_shorthand)

```


## Time series modeling

Ideas:

- Cross-correlation analysis (ccf) 
- Granger causality
- Changepoint / onset analysis


CCF: Here, we quantify the correlation between each AC and FB at various "lags". 

```{r}

### Multiple obs. per step/token seen ,so group across?
df_summ_fb_tokens = df_summ_fb %>%
  filter(stage == "stage1") %>%
  group_by(model_shorthand, tokens_seen_numeric_mod, step) %>%
  summarise(fb_accuracy = mean(mean_accuracy))

### Multiple obs. per step/token seen ,so group across?
df_all = df_summ_attn %>%
  mutate(model_shorthand = str_to_title(model_shorthand)) %>%
  filter(model_shorthand == "Olmo 2 13b") %>%
  transmute(ac_accuracy = mean_accuracy) %>%
  group_by(model_shorthand, tokens_seen_numeric_mod, step, question_id) %>%
  summarise(ac_accuracy = mean(ac_accuracy)) %>%
  # filter(question_id == 1) %>%
  inner_join(df_summ_fb_tokens) %>%
  arrange(tokens_seen_numeric_mod)


### New plot
plot_data <- df_all %>%
  dplyr::select(model_shorthand, tokens_seen_numeric_mod, step, question_id, 
         ac_accuracy, fb_accuracy) %>%
  pivot_longer(cols = c(ac_accuracy, fb_accuracy),
               names_to = "task_type",
               values_to = "accuracy") %>%
  mutate(task_type = case_when(
    task_type == "ac_accuracy" ~ "Situation Model",
    task_type == "fb_accuracy" ~ "False Belief"
  )) %>%
  mutate(q_label = case_when(
    question_id == 1 ~ "Item was first in {START}",
    question_id == 2 ~ "At end of story, item in {END}",
    question_id == 3 ~ "Original person was {X}",
    question_id == 4 ~ "Second person was {Y}"
  ))


# Create the plot
plot_data %>%
  filter(question_id <= 2) %>%
  ggplot(aes(x = tokens_seen_numeric_mod, 
                      y = accuracy, 
                      color = task_type)) +
  geom_line(size = 1, alpha = 0.8) +
  geom_point(size = 2, alpha = 0.6) +
  facet_wrap(~ reorder(q_label, question_id), ncol = 2) +
  scale_x_log10() +
  # scale_x_continuous(labels = scales::scientific,
  #                   name = "Tokens Seen") +
  scale_y_continuous(limits = c(0, 1),
                     name = "Accuracy") +
  scale_color_manual(values = c("Situation Model" = "#808080", 
                                 "False Belief" = "#21908CFF")) +
  labs(x = "Tokens Seen (Log10)",
       color = "Task Type") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold", size = 12))



plot_data %>%
  filter(question_id > 2) %>%
  ggplot(aes(x = tokens_seen_numeric_mod, 
                      y = accuracy, 
                      color = task_type)) +
  geom_line(size = 1, alpha = 0.8) +
  geom_point(size = 2, alpha = 0.6) +
  facet_wrap(~ reorder(q_label, question_id), ncol = 2) +
  scale_x_log10() +
  # scale_x_continuous(labels = scales::scientific,
  #                   name = "Tokens Seen") +
  scale_y_continuous(limits = c(0, 1),
                     name = "Accuracy") +
  scale_color_manual(values = c("Situation Model" = "#808080", 
                                 "False Belief" = "#21908CFF")) +
  labs(x = "Tokens Seen (Log10)",
       color = "Task Type") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold", size = 12))



plot_data %>%
  ggplot(aes(x = tokens_seen_numeric_mod, 
                      y = accuracy, 
                      color = task_type)) +
  geom_line(size = 1, alpha = 0.8) +
  geom_point(size = 2, alpha = 0.6) +
  facet_wrap(~ reorder(q_label, question_id),, ncol = 2) +
  scale_x_log10() +
  # scale_x_continuous(labels = scales::scientific,
  #                   name = "Tokens Seen") +
  scale_y_continuous(limits = c(0, 1),
                     name = "Accuracy") +
  scale_color_manual(values = c("Situation Model" = "#808080", 
                                 "False Belief" = "#21908CFF")) +
  labs(x = "Tokens Seen (Log10)",
       color = "Task Type") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold", size = 12))




model_data_wide <- df_all %>%
  pivot_wider(
    names_from = question_id,
    values_from = c(ac_accuracy),  # or just the columns you want
    names_prefix = "q"  # optional: adds "q" prefix to column names
  )


# Compute cross-correlation
ccf_result <- ccf(
  model_data_wide$q1, 
  model_data_wide$fb_accuracy,
  lag.max = 10,  # test up to 10 lags
  plot = TRUE,
  main = "Cross-Correlation: Situation Model → False Belief"
)

# Positive lags indicate situation model leads FB
print(ccf_result)

max_lag <- ccf_result$lag[which.max(ccf_result$acf)]
max_corr <- max(ccf_result$acf)

cat(sprintf("Maximum correlation: %.3f at lag %d\n", max_corr, max_lag))
cat(sprintf("This means q1 leads FB by %d time steps\n", abs(max_lag)))


# Compute cross-correlation
ccf_result <- ccf(
  model_data_wide$q2, 
  model_data_wide$fb_accuracy,
  lag.max = 10,  # test up to 10 lags
  plot = TRUE,
  main = "Cross-Correlation: Situation Model → False Belief"
)

# Positive lags indicate situation model leads FB
print(ccf_result)

max_lag <- ccf_result$lag[which.max(ccf_result$acf)]
max_corr <- max(ccf_result$acf)

cat(sprintf("Maximum correlation: %.3f at lag %d\n", max_corr, max_lag))
cat(sprintf("This means q1 leads FB by %d time steps\n", abs(max_lag)))

# Compute cross-correlation
ccf_result <- ccf(
  model_data_wide$q3, 
  model_data_wide$fb_accuracy,
  lag.max = 10,  # test up to 10 lags
  plot = TRUE,
  main = "Cross-Correlation: Situation Model → False Belief"
)

# Positive lags indicate situation model leads FB
print(ccf_result)

max_lag <- ccf_result$lag[which.max(ccf_result$acf)]
max_corr <- max(ccf_result$acf)

cat(sprintf("Maximum correlation: %.3f at lag %d\n", max_corr, max_lag))
cat(sprintf("This means q1 leads FB by %d time steps\n", abs(max_lag)))


# Compute cross-correlation
ccf_result <- ccf(
  model_data_wide$q4, 
  model_data_wide$fb_accuracy,
  lag.max = 10,  # test up to 10 lags
  plot = TRUE,
  main = "Cross-Correlation: Situation Model → False Belief"
)

# Positive lags indicate situation model leads FB
print(ccf_result)

max_lag <- ccf_result$lag[which.max(ccf_result$acf)]
max_corr <- max(ccf_result$acf)

cat(sprintf("Maximum correlation: %.3f at lag %d\n", max_corr, max_lag))
cat(sprintf("This means q1 leads FB by %d time steps\n", abs(max_lag)))
```



Granger causality:

```{r}
library(lmtest)
library(vars)

# Prepare time series
ts_data <- model_data_wide %>%
  dplyr::select(tokens_seen_numeric_mod, q1, q2, q3, q4, fb_accuracy) %>%
  drop_na()

# Prepare data as multivariate time series
ts_data_clean <- model_data_wide %>%
  ungroup() %>%
  dplyr::select(q1, fb_accuracy) %>%
  drop_na()

# Test different lag orders
VARselect(ts_data_clean, lag.max = 10, type = "const")

# Test if situation model "Granger-causes" false belief
grangertest(fb_accuracy ~ q1, 
            order = 1,  # number of lags
            data = ts_data)

grangertest(fb_accuracy ~ q2, 
            order = 3,  # number of lags
            data = ts_data)

```
